---
title: "Workshop 8 TEMPLATE"
subtitle: "12pm"
author: "Niki van der Poel"
date: "May 22 2026"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest"
)

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating new object for clean data
drought_exp_clean <- drought_exp |> 
  # cleaning names to make legible
  # adding underscores and replaces capitals
  clean_names() |> 
  # creating new column with scientific name
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # Creating new column with treatment names
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # reordering species' names
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
# reordering treatment levels  
mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting certain columns to keep/ making data concise
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  





Insert code to display 10 random rows from the data frame here.  


```{r}
#| label: ten rows

slice_sample(
  drought_exp_clean, # clean data fram to display
  n = 10 # show ten rows
)

```



# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 

```{r}
#| label: making a scatterplot

#ggplot base layer, drought clean
ggplot(data = drought_exp_clean,
      # x axis = sla
      # y axis= total mass
        mapping = aes(x = sla,
                     y = total_g)
) +
  #adding points
  geom_point(size = 2) +
  # labelling axes with units and adding a title
  labs( y = "Total plant mass (g)",
        x = "Specific Leaf Area (SLA) (cm"^2~"/g)",
    title = "As SLA increases, total plant mass increases")
```


## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.


```{r}
#| label: making a jitter to compare water treatments
#| fig-height: 6
#| fig-width: 9

ggplot(data = drought_exp_clean,
      # x axis = water treatment
      # y axis= total mass
        mapping = aes(x = water_treatment,
                     y = total_g,
                     color = water_treatment)
) +
  #adding points
  geom_jitter(height = 0, 
              width = 0.1
              ) +
  scale_color_manual(values = c(
    "Drought stressed" = ds_col,
    "Well watered" = ww_col
  )) +
  # labelling axes with units and adding a title
  labs( y = "Total plant mass (g)",
        x = "Water Treatment",
    title = "Well-watered individuals tend to have higher mass than drought stressed") +
  theme(legend.position = "none")
```





## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.

```{r}
#| label: making a jitter to compare species
#| fig-height: 6
#| fig-width: 8

ggplot(data = drought_exp_clean,
      # x axis = species
      # y axis= total mass
        mapping = aes(x = species_name,
                     y = total_g,
                     color = species_name)
) +
  #adding points
  geom_jitter(height = 0, 
              width = 0.1
              ) +
  #wrapping x axis names
  scale_x_discrete(labels = label_wrap(10)) +
  # labelling axes with units and adding a title
  labs( y = "Total plant mass (g)",
        x = "Species name",
    title = "Species differ in plant mass") +
  #custom colors
  scale_color_manual(
    values = c(
      "Lotus scoparius" = lotsco_col,
      "Penstemon centranthifolius" = pencen_col,
      "Salvia leucophylla" = salleu_col,
      "Encelia californica" = enccal_col,
      "Stipa pulchra" = stipul_col,
      "Eschscholzia californica" = esccal_col,
      "Grindelia camporum" = gricam_col
    )
  ) +
  theme(legend.position = "none")
```

# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model

model0 <- lm(
total_g ~ 1,
data = drought_exp_clean
)

# model 1: all predictors

model1 <- lm(
total_g ~ sla + water_treatment + species_name,
data = drought_exp_clean
)

# model 2: SLA and water treatment

model2 <- lm(
total_g ~ sla + water_treatment,
data = drought_exp_clean
)

# model 3: SLA and species

model3 <- lm(
total_g ~ sla + species_name,
data = drought_exp_clean
)

# model 4: water treatment and species

model4 <- lm(
total_g ~ species_name + water_treatment,
data = drought_exp_clean
)

# model 5: SLA

model5 <- lm(
total_g ~ sla,
data = drought_exp_clean
)

# model 6: water treatment

model6 <- lm(
total_g ~ water_treatment,
data = drought_exp_clean
)

# model 7: species

model7 <- lm(
total_g ~ species_name,
data = drought_exp_clean
)

```

## Model diagnostics

```{r}
#| label: checking-diagnostics
#| fig-height: 6
#| fig-width: 6

#displaying plots
par(mfrow = c(2,2))
#plotting diagnostics
plot(model4)
```

- residuals are normally distrubted (QQ plot)
- residuals are homoscedastic (residuals and scal-location)
- no outliers that could be influencing model coefficient estimates (leverage)

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: model-selection

AICc(
model0,
model1,
model2,
model3,
model4,
model5,
model6,
model7
) |> 
arrange(AICc)
```



Insert code to display the summary for the best model. 

```{r}
#| label: model summary

```




## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors

**How much variation in the response (total mass, in grams) does this model explain?**

72.6% of variation in plant mass is explained by this model

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

- Species influences plant mass
- water treatment influences plant mass

- '(Intercept)': The intercept is the model prediction for plant mass for Lotus scoparius in the drought stressed treatment (lotus scoparius and drought stressed were the reference levels in the summary)
- 'water_treatmentWell waters': model prediction for dif in lotus scoparius mass between well watered and drought stressed treatments.
- 'species_namePenstemon centranthifolius': model prediction for dif in penstemon centranthifolius mass between well watered and drought stressed treatments

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: model-predition

preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |> 
  rename(water_treatment = x,
         species_name = group)
  
```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
#| label: final figures
#| fig-height: 6
#| fig-width: 10


ggplot(data = drought_exp_clean,
       mapping = aes(x = water_treatment,
                     y = total_g,
                     color = water_treatment)) +
  # individual plants for each treatment
  geom_jitter(height = 0,
              width = 0.1,
              alpha = 0.4,
              shape = 21) +
  # predictions and confidence intervals
  geom_pointrange(data = preds,
                  mapping = aes(x = water_treatment,
                                y = predicted,
                                ymin = conf.low,
                                ymax = conf.high)) +
  #creates different panels for each series
  facet_wrap(~ species_name
            ) +
  # custom colors
  scale_color_manual(values = c(
    "Drought stressed" = ds_col,
    "Well watered" = ww_col
  )) +
  theme(legend.position = "none") +
  labs(x = "Water Treatment",
       y = "Total plant mass (g)")

```


# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







